---
title: "MorphoGAM: Detect spatially variable genes by projecting to morphologically relevant curves"
author: "R package version 1.1.0"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## System Requirements

`R` is required to use `MorphoGAM`. In development, `R` version 4.3.0 and greater were used, but there may be compatibility with previous versions. 

## Installation 

From the R console, `devtools::install_github("phillipnicol/MorphoGAM")`. Installation should take less than a minute on a standard machine. 

## Demo 

### Step 1: Estimate morphologically relevant coordinates

The first step in running `MorphoGAM` is to define a curve from which the morphologically relevant coordinates are defined. To demonstrate this, we use the swiss roll example: 

```{r, warning=FALSE, message=FALSE}
set.seed(1)
library(MorphoGAM)
library(tidyverse)
xy <- MorphoGAM:::makeSwissRoll()
data.frame(x=xy[,1],y=xy[,2]) |> ggplot(aes(x=x,y=y)) + 
  geom_point(size=0.5) + theme_bw()
```

The function `CurveFinder()` applies the automatic curve estimation method 

```{r}
fit <- CurveFinder(xy)
```


The `fit` object contains plots of the first two morphologically relevant coordinates and the fitted curve: 

```{r}
fit$curve.plot
fit$coordinate.plot #First morphologically relevant coordinate 
fit$residuals.plot #Second morphologically relevant coordinate
```

The morphologically relevant coordintes can be accessed via `fit$xyt`: 

```{r}
fit$xyt |> head()
```


In some cases the user may wish to draw the curve by hand. For this we provide an interactive shiny app that can be run locally using the function `CurveFinderInteractive()`. Once the app is running you can click the sequence of points defining the curve, then press the "Smooth" button to fit the curve. Once the smoothing is done the app can be closed and `fit` will be returned.

```{r, eval=FALSE}
  #Running this opens a shiny app
  fit <- CurveFinderInteractive(xy)
```

### Step 2: Apply GAM to morphologically relevant coordinates

The next step is to identify genes with variable expression along the curve (or in the orthogonal direction). Here we generate a synthetic count matrix `Y` with one spatially interesting gene: 

```{r}
Y <- matrix(rpois(100*nrow(xy), lambda=1),
            nrow=100, ncol=nrow(xy))

eta <- -3*fit$xyt$t + 2
Y[1,] <- rpois(nrow(xy),lambda=exp(eta))

rownames(Y) <- paste("Gene", 1:nrow(Y))
```

Now we apply the generalized additive model (GAM): 

```{r}
mgam <- MorphoGAM(Y, curve.fit=fit,
                  design = y ~ s(t, bs="cr"))
```

The `bs = "cr"` specifies cubic regression splines in the GAM, although this can be modified to periodic splines or other basis functions provided by `mgcv`. We may wish to sort the results matrix to rank genes by summaries of the estimated function:

```{r}
mgam$results |> arrange(desc(peak.t)) |> head()
```

The results indicate gene $1$ has a significant peak and range (region of increased expression), and we can visually confirm this by using `plotGAMestimate()` to plot the entire function:

```{r}
plotGAMestimates(Y,genes=c("Gene 1", "Gene 2"),
                 mgam_object = mgam,
                 curve_fit=fit,
                 type="t")
```


To also identify genes that vary in the direction of the second morphologically relevant coordinate, add the term `s(r, ...)` to the `design` argument in `MorphoGAM`. 

```{r, eval=FALSE}
mgam_with_r <- MorphoGAM(Y, curve.fit=fit,
                          design = y ~ s(t, bs="cr") + s(r, bs="cr"))
```



## Reference

If you use `MorphoGAM` in your work, please cite:


