---
title: "MorphoGAM: Detect spatially variable genes by projecting to morphologically relevant curves"
author: "R package version 1.0.0"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## System Requirements

`R` is required to use `MorphoGAM`. In development, `R` version 4.3.0 and greater were used, but there may be compatibility with previous versions. 

## Installation 

From the R console, `devtools::install_github("phillipnicol/MorphoGAM")`. Installation should take less than a minute on a standard machine. 

## Demo 

The first step in running `MorphoGAM` is to define a curve from which the morphologically relevant coordinates are defined. To demonstrate this, we use the swiss roll example: 

```{r, warning=FALSE, message=FALSE}
set.seed(1)
library(MorphoGAM)
library(tidyverse)
xy <- MorphoGAM:::makeSwissRoll()
data.frame(x=xy[,1],y=xy[,2]) |> ggplot(aes(x=x,y=y)) + 
  geom_point(size=0.5) + theme_bw()
```

The function `CurveFinder()` applies the automatic curve estimation method 

```{r}
fit <- CurveFinder(xy)
```


The `fit` object contains plots of the first two morphologically relevant coordinates and the fitted curve: 

```{r}
fit$curve.plot
fit$coordinate.plot
fit$residuals.plot
```

To interactively specify the curve, use the following function. Press the "Smooth" button when the points have been specified and then close the app to obtain the result: 

```{r, eval=FALSE}
  #Running this opens a shiny app
  fit <- CurveFinderInteractive(xy)
```

The next step is to identify genes with variable expression along the curve (or in the orthogonal direction). Here we generate a synthetic count matrix `Y`: 

```{r}
Y <- matrix(rpois(100*nrow(xy), lambda=1),
            nrow=100, ncol=nrow(xy))

eta <- -3*fit$xyt$t + 2
Y[1,] <- rpois(nrow(xy),lambda=exp(eta))

rownames(Y) <- paste("Gene", 1:nrow(Y))
```

Now we apply the generalized additive model (GAM): 

```{r}
mgam <- MorphoGAM(Y, curve.fit=fit,
                  design = y ~ s(t, bs="cr"))
```

The `bs = "cr"` specifies cubic regression splines in the GAM, although this can be modified to periodic splines or other basis functions provided by `mgcv`. We may wish to sort the results matrix to rank genes by summaries of the estimated function:

```{r}
mgam$results |> arrange(desc(peak.t)) |> head()
```

The results indicate gene $1$ has a significant peak and range (region of increased expression), and we can visually confirm this by using `plotGAMestimate()` to plot the entire function:

```{r}
plotGAMestimates(Y,genes=c("Gene 1", "Gene 2"),
                 mgam_object = mgam,
                 curve_fit=fit,
                 type="t")
```


To also identify genes that vary in the direction of the second morphologically relevant coordinate, add the term `s(r, ...)` to the `design` argument in `MorphoGAM`. 

```{r, eval=FALSE}
mgam_with_r <- MorphoGAM(Y, curve.fit=fit,
                          design = y ~ s(t, bs="cr") + s(r, bs="cr"))
```



## Reference

If you use `MorphoGAM` in your work, please cite:


