---
title: "MERFISH Mouse Mucosa"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**TODO:** Data availability. This example assumes that the mucosa coordinates and count matrix have
already been saved locally. The coordinate file should contain one row per
cell and two spatial coordinate columns. The count file should contain genes
as rows and cells as columns.

```{r setup, message = FALSE, warning = FALSE}
library(MorphoGAM)
library(tidyverse)
```

```{r load-data}
# Replace these with the local paths where the data have been saved.
resolve_data_path <- function(path) {
  candidates <- c(
    path,
    file.path("..", "..", path)
  )
  match <- candidates[file.exists(candidates)][1]
  if (is.na(match)) {
    stop("Could not find data file: ", path)
  }
  match
}

coords_file <- resolve_data_path("analysis/data/mucosa_sub_coords.csv")
counts_file <- resolve_data_path("analysis/data/counts_sub_mucosa.csv")

xy <- read.csv(coords_file, row.names = 1, check.names = FALSE)
xy <- as.matrix(xy)
storage.mode(xy) <- "double"

Y <- read.csv(counts_file, row.names = 1, check.names = FALSE)
Y <- as.matrix(Y)
storage.mode(Y) <- "double"
```

```{r plot-coordinates}
xy_df <- as.data.frame(xy)
colnames(xy_df) <- c("x", "y")

ggplot(xy_df, aes(x = x, y = y)) +
  geom_point(size = 0.1) +
  coord_fixed() +
  theme_bw()
```

The mucosa in this example is approximately loop-shaped, so we ask
`CurveFinder()` to estimate a loop and use a smaller neighborhood size.

```{r curvefinder}
curve_fit <- MorphoGAM::CurveFinder(
  xy,
  knn = 14, #Value used in paper, use k = "auto" for automatic
  loop = TRUE
)

#Plot the fit and two coordaintes
curve_fit$curve.plot
curve_fit$coordinate.plot
curve_fit$residuals.plot
```

Next, we fit a smooth cyclic effect along the loop coordinate `t` and a radial
smooth effect for distance from the curve.

```{r morphogam}
mgam <- MorphoGAM::MorphoGAM(
  Y,
  curve.fit = curve_fit,
  design = y ~ s(t, bs = "cc") + s(r, bs = "cr")
)
```

Genes with large `peak.t` values have a strong fitted change along the loop
coordinate.

```{r plot-top-t}
top_t_genes <- rownames(mgam$results)[
  order(mgam$results$peak.t, decreasing = TRUE)[1:6]
]

MorphoGAM::plotGAMestimates(
  Y,
  genes = top_t_genes,
  curve_fit = curve_fit,
  mgam_object = mgam,
  nrow = 2
)
```

We can similarly inspect genes with large radial variation.

```{r plot-top-r}
top_r_genes <- rownames(mgam$results)[
  order(mgam$results$range.r, decreasing = TRUE)[1:6]
]

MorphoGAM::plotGAMestimates(
  Y,
  genes = top_r_genes,
  curve_fit = curve_fit,
  mgam_object = mgam,
  type = "r",
  nrow = 2
)
```

Now we can make the functional PCA plot to see
the dominant modes of spatial variation:

```{r functional-pca}
MorphoGAM::plotFPCloading(
  mgam,
  curve.fit = curve_fit,
  type = "t",
  L=1 #First component
)

MorphoGAM::plotFPCloading(
  mgam,
  curve.fit = curve_fit,
  type = "t",
  L=2 #Second component
)
```
